Active Code Completion

### 2012-03-18 vlhcc11/vlhcc11.tex
Initial paper move.

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
\texttt{\url{https://github.com/cyrus-/graphite}}
}

@@

% make the title area
% \maketitle

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
% \IEEEpeerreviewmaketitle

\section{Developer Survey}

Prior to implementation, we conducted a large online survey of professional software developers in order to extract criteria that govern whether our system would be useful to them. More specifically, we sought to address the following questions:

\begin{itemize}
\item What are specific use cases for active code completion in a professional setting? 
\item What are some functional criteria that predict whether a class would benefit from active code completion?
\item What are some usability criteria that should govern palette interface designs?
\item Which capabilities must a active code completion architecture support to enable these use cases and designs?
\end{itemize}

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mock-up screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.

We posted a link to our survey on a popular  collaborative filtering website \cite{mehta_robust_2007} called {\tt /r/programming} on reddit.com\footnote{\texttt{http://www.reddit.com/r/programming}}. At the time of our study, the website listed approximately 340,000, though members were not required to registered to read the content,  and contained content that could be considered relevant to our intended audience. An additional 22 participants were recruited using a mass email to local computer science graduate students. The recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.

Participants were asked to take an online survey that would take approximately 20 minutes to complete. Of the 696 people who started the survey by answering the first question, 473 participants completed it. Of these, 16 participants were computer science graduate students at Carnegie Mellon and 457 were from \texttt{/r/programming}.

Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems. Many of these findings can also be generalized to constrain the design of visual programming languages, particularly those that aim be useful to professional developers, rather than novice programmers (e.g., Barista \cite{ko_barista:_2006}). We summarize the survey questions and the results in the following subsections.
% In fact, active code completion can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional representations of program logic. This may help address some of the usability challenges previously associated with visual languages (cf. [?]).

\subsection{Familiarity with Programming Languages and Editors}

We first asked their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''. 61.1\% of the respondents expressed themselves as experts on at least one language, and 35.7\% were very familiar with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

We also asked participants to select the integrated development environments (IDEs) and code editors that they were familiar with. The Eclipse IDE was familiar to 87.1\% of participants. This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%. Participants could also enter ``other'' choices. A number of editors and IDEs were entered, including Xcode, Textmate and Notepad++. None of these were dominant.

\subsection{Mock-ups}

\begin{figure}
  \includegraphics[width=\linewidth]{mockup-palette-small.png}
  \caption{An example mock-up palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
  \label{regex}
\end{figure}

Next, we presented participants with a series of mock-up palettes for a Color class, a regular expression class (Fig. \ref{regex}), and a SQL query class. Participants were also shown a mock-up demonstrating how a user could invoke the palette, and a mock-up showing the code that would be inserted once a selection had been made.

Before presenting each mock-up, we gathered information about the user's familiarity with the class and the strategies that they would likely use to instantiate the class under normal circumstances. For the Color class, the majority of participants indicated that they would look in the code completions menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant. Another 14.0\% indicated that they would use an external tool to determine the RGB values. Few participants were unfamiliar with regular expressions and no participants were unfamiliar with SQL queries which provides further evidence that our participants were not novice developers. Fig. \ref{strategies} summarizes their usual strategies for instantiating regular expressions and SQL queries.

\begin{figure}
\begin{tabular}{lccc}
 & Regular Expressions &  & SQL\\
 \hline
Separate test script & 29.6\% &  & 15.4\%\\
Guess and check & 14.0\% &  & 16.1\%\\
External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
Search for examples & 12.3\% & & 5.1\% \\
Other & 6.2\% & & 4.9\% \\
\hline
\end{tabular}
\caption{Typical strategies for regular expressions and SQL queries.}
\label{strategies}
\end{figure}

\begin{figure}
\begin{tabular}{crccccc}\\\\
\textsc{class}
& 
& \begin{rotate}{20}Nearly every time\end{rotate}
& \begin{rotate}{20}Most of the time\end{rotate}
& \begin{rotate}{20}Some of the time\end{rotate}
& \begin{rotate}{20}Rarely\end{rotate}
& \begin{turn}{20}Never\end{turn}\\
\hline
Color &\vline& 9.6\% & 22.1\% & \textbf{32.4\%} & 28.2\% & 7.7\%\\
RegExp &\vline& \textbf{36.6\%} & 29.5\% & 21.8\% & 7.3\% & 4.8\%\\
SQL & \vline &18.2\% & 19.3\% & \textbf{30.9\%} & 20.4\% & 11.4\%\\
\hline
\end{tabular}
\caption{The distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class. What portion of the time, in these situations, do you think you would use this feature?''}
\label{frequencies}
\end{figure}

Participants were positive about our mock-ups. We asked the users to rate the usefulness of each mock-up palette, and their responses are summarized in Fig. \ref{frequencies}. For each palette, more than half of the respondents expressed that they would use it at least some of the time. Especially, people liked the regular expression palette a lot.

\section{Design Implications}

We analyzed the open-ended responses to extract design principles and constraints. We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette. Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes. The extracted principles and constraints are summarized in the following subsections. The parenthesized number following each design constraint indicates how many people commented on that issue.

\subsection{System Design Constraints}

System design constraints are the criteria that should be taken into account in the design of the overall system architecture.

\subsubsection{Handling Separation of Concerns (183?)}

Participants were very wary of including constant data into the source code directly. Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file. Sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette. Interestingly, most people seemed to consider regular expressions as part of the program logic. There were few people who complained about including the regular expression pattern string in the source code as opposed to the other two classes.

%Examples of feedbacks in this category include:
%
%\begin{itemize}
%	\item The use of magic numbers in the source code is a bad design (54)
%	\item Color picking is the `designer's job' (33)
%	\item Would only use to define named constants (17)
%	\item Would rather use color scheme / theme (3)
%	\item Would use Object-Relational Mapping (ORM) / Language INtegrated Query (LINQ) instead of raw SQL query string (27 of 142)
%	\item These palettes might be useful for prototyping, but not for real production (14)
%	\item Instead of including the test cases into comments, they should be stored as unit tests (35 of 143)
%\end{itemize}

\subsubsection{Reversibility (19?)}

Some participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.

\subsubsection{Palette Settings and State (41?)}

Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden. % The followings are the comments that are related to this category.

%\begin{itemize}
%	\item Want recently used regular expressions, and control over the comments (12)
%	\item Persistent database connection information is needed / want to manage connection pools (9)
%	\item Want recent colors, capability of setting favorite colors (20)
%\end{itemize}

\subsubsection{Interaction with Code Context (13?)}

Some people wanted to interact with the code context. For example, people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.

\subsubsection{Performance (?)}

Developers were concerned about the performance of these palettes. Some of the participants said that they would use these palettes only if it does not slow down the IDE. Also, this was the most popular comment on the reddit thread.

\subsubsection{IDE Independence (?)}

Several participants expressed a desire for IDE or even language independence for these palettes.

\subsection{General Palette Design Considerations}

General palette design considerations are about how to make more usable palettes.

\subsubsection{Simplicity vs. Capability}

Even though the participants were shown the same mockup screenshots, their reactions were split. Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names. The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively). They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.

\subsubsection{Keyboard Navigability (12)}

12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.

\subsection{Suggestions from the Participants}
At the end of the survey, we solicited suggestions from the participants about what types of classes could benefit from the active code completion palettes. 119 participants suggested various types of classes and we classified the suggestions into several categories.

\subsubsection{Alternative/Tricky/Literal Syntax (16)}
These are the classes of which syntax is tricky, or complex. Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping. We address this issue in more detail in Section \#.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item String
%		\item Collections (e.g., dictionary)
%		\item Vector, Matrix
%		\item Embedded languages
%		\item URLs, Paths
%	\end{itemize}
%	
%	\item Expected features
%	
%	\begin{itemize}
%		\item Proper escaping
%		\item Syntax highlighting
%		\item Auto-completion
%	\end{itemize}
%\end{itemize}

\subsubsection{Unclear Parameter Implications (11)}
The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications. Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly. Also, the palette can be a control panel which facilitates modifying the parameters as in Juxtapose [?].

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Audio tweaks (modifying pitch, volume, etc.)
%		\item 3D transformation matrices
%		\item number / string / date formatting
%		\item message box / input box (e.g., \texttt{JOptionPane})
%	\end{itemize}
%\end{itemize}

\subsubsection{Query Languages (17)}
Similar to the regular expression and SQL palette we presented, the developers also wanted to check if the other types of query strings (e.g., XPath / XQuery) are correct or not by checking the results directly on the palette.

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Regular expression
%		\item SQL query
%		\item XPath / XQuery
%	\end{itemize}
%\end{itemize}

\subsubsection{Graphical Elements (27)}
The other popular category was graphical elements. We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others. Also, people wanted to check or directly manipulate the GUI frames and layouts.

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Color, Shape, Brush, Font, etc.
%		\item JFrame, Layouts, any swing components
%	\end{itemize}
%\end{itemize}

\subsection{Complex Instantiation Procedures (5)}
Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code. For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class. It is error prone to use this class because it requires try/catch block, and also one must close the file after reading it. By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects. In the same way, it can alleviate the factory pattern usability problem \cite{ellis_factory_2007}. As long as the developers remember which class to use, they will not need to remember how to instantiate that class.

%\begin{verbatim}
%BufferedReader reader = null;
%try {
%    reader = new BufferedReader(
%        new FileReader(filePath));
%} catch (IOException e) {
%    e.printStackTrace();
%}
%// use reader here
%reader.close();
%\end{verbatim}

\subsection{Describe by Example (2)}
Sometimes, it is possible to describe an object by examples. For instance, if there is a class which represents a shortcut key combination, we can easily instantiate an object of that class by pressing the actual shortcut key on the interactive palette.

%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item Keyboard Keys
%		\item Regular expression
%	\end{itemize}
%\end{itemize}

\subsection{Integrating with Documentation, Tutorial (7)}
Some participants suggested integrating the documentation or the tutorials into the interactive palettes. If the developers are not familiar of the specific API class, the palette could tell them how to use it.

% conference papers do not normally have an appendix

% use section* for acknowledgement
%\section*{Acknowledgment}

%The authors would like to thank...

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
\texttt{\url{https://github.com/cyrus~/graphite}}
}

@@

% make the title area
\maketitle

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}

Software developers today make heavy use of the code completion features available in modern code editors \cite{murphy_how_2006}. By navigating and selecting from a floating menu containing the names of variables, fields, methods, types and code snippets, a developer can avoid many common spelling and logic errors, avoid unnecessary keystrokes and explore unfamiliar APIs without leaving the editor window. To ensure that the items featured in this menu are relevant, the editor conducts a static analysis of the surrounding code context. %In many editors, an area adjacent to this menu displays documentation associated with the  currently highlighted item.

Several refinements and additions to the code completion menu have been previously suggested in the literature. These have focused on using additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured items. Empirical evidence presented in these studies suggests that these enhancements further improve developer productivity.

In this paper, we propose a complementary technique called {\it active code completion}. When the developer invokes the code completion menu, the editor looks for a {\it palette definition} associated with the type of the expression being entered. If found, an option to use this palette is added to the code completion menu.
When the developer selects this option, source code is not inserted immediately. Instead, the palette definition takes control of the code completion interface. The developer can then interact with this interface to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the object's behavior. When the developer indicates that she is satisfied with these choices, the palette generates code that is inserted at the cursor.

Before designing and implementing such a system, we sought to address the following questions:

\begin{itemize}
\item What are some specific use cases for active code completion in a professional development setting? 
\item Which functional criteria are common to types that would benefit from an associated palette?
\item What usability criteria should inform palette interface designs in this context?
\item Which capabilities must the active code completion architecture have to enable these use cases and designs?
\end{itemize}

To answer these questions, we began by conducting a large online survey of professional developers. Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems.

Based on this information, we designed and implemented an Eclipse plug-in that provides active code completion for the Java language, and implemented some example palettes atop this system. 
Finally, we evaluated the usefulness of one such palette, designed to assist developers as they write regular expressions, with a controlled lab study. The data and observations from this study provide empirical evidence in support of the view that an active code completion system would be useful to professional developers.

\section{Developer Survey}
The subject pool for the preliminary survey consisted primarily of visitors to a large programming-oriented 
collaborative filtering forum\footnote{http://www.reddit.com/r/programming}.
A small number of participants were also recruited using a
mass email to local computer science graduate students. Participants were asked to take an online survey taking  approximately 20 minutes to complete and background information collected from the 475 participants indicated that the overwhelming majority of participants were professional developers.

Participants were shown three mockup palettes, along with mockups of the invocation procedure described above, for classes representing colors, regular expressions and SQL queries. They were asked to indicate whether they would find these palettes useful. Their responses are summarized in Figure 1 and indicate that these developers found the concept potentially useful, with a particular preference for the palette shown for regular expressions.

Participants were also asked to provide free-form comments on each palette. These responses revealed a number of concerns about this system. These include issues related to handling separation of concerns, reinvocation, palette settings and state, interactions between the palette and the code context, language and editor independence, and keyboard navigability. These concerns strongly influenced the design of the code completion system described in the next section.

Finally, we solicited suggestions for classes that may benefit from an associated palette. We received a number of interesting suggestions, which we broadly classified into categories. Examples include classes that would benefit from an alternative syntax (e.g. dictionaries), where the implications of a parameter choice are difficult to predict (e.g. 3D transformation matrices), where the instantiation procedure is non-trivial (e.g. factory methods) and where parameters could be provided by example (e.g. key combinations).
\begin{figure*}\begin{center}
  \includegraphics[scale=.4]{regex.png}\end{center}
  %\caption{Test}
  \vspace{-15pt}
  \end{figure*}

\begin{figure}
\begin{tabular}{crccccc}\\\\
\textsc{class}
& 
& \begin{rotate}{20}Nearly every time\end{rotate}
& \begin{rotate}{20}Most of the time\end{rotate}
& \begin{rotate}{20}Some of the time\end{rotate}
& \begin{rotate}{20}Rarely\end{rotate}
& \begin{turn}{20}Never\end{turn}\\
\hline
Color &\vline& 9.6\% & 22.1\% & \textbf{32.4\%} & 28.2\% & 7.7\%\\
RegExp &\vline& \textbf{36.6\%} & 29.5\% & 21.8\% & 7.3\% & 4.8\%\\
SQL & \vline &18.2\% & 19.3\% & \textbf{30.9\%} & 20.4\% & 11.4\%\\
\hline
\end{tabular}
\caption{The distribution of responses to the question: ``Consider situations where you need to instantiate the [specified] class. What portion of the time, in these situations, do you think you would use this feature?''}
  \vspace{-15pt}

\end{figure}
\section{Prototype Implementation}

Next, we implemented an active code completion system as an Eclipse Java plug-in called \textsc{Graphite}, an acronym for \textbf{Gra}phical \textbf{P}alettes \textbf{H}elp \textbf{I}nstantiate \textbf{T}ypes in the \textbf{E}ditor.

\subsection{Association Model}
Graphite provides two mechanisms to associate a palette with a class. The {\it annotation-based association model} allows the developer of a class to associate a palette definition with it using a Java annotation, \texttt{GraphitePalette}. The annotation contains the URL of the palette and metadata used to control the description of the palette in the code completion menu. This association model is beneficial because end-users need not realize that a palette is available for a class -- when they invoke the code completion menu, they {\it discover} that a palette exists. This stands in contrast to relevant external tools, which must be explicitly discovered by users.

For classes which are not amenable to direct modification, such as those in the Java standard library, an {\it external association model} is available in the Graphite preferences pane. Users can explicitly associate a palette with the fully-qualified name of a class using this model.

\subsection{Palette Implementation Model}

A common concern expressed in our developer survey was that active code completion should not rely on a particular editor environment. To satisfy this constraint, we could not use an IDE-specific user interface framework (e.g. SWT and JFace for Eclipse). Instead, we designed Graphite so that palettes were written using HTML and Javascript. Our preliminary study confirmed the fact that these languages are well-known amongst Java developers. A small Javascript bridge API was developed to allow palettes to access the current selection in the editor (used to allow reinvocation of palettes) and to insert code at the cursor. The Eclipse plug-in simply displays the palette in a floating browser window, without any associated chrome, and responds to invocations of bridge functions.

This strategy could be straightforwardly replicated in other editor environments, without requiring that palette definitions be modified. Indeed, with appropriate logic in the bridge API, one palette may be able to support several distinct languages.

\section{Lab Study}
%\begin{figure*}[ht]
%  \caption{test}
%\end{figure*}

In order to evaluate the usefulness and usability of Graphite, we conducted a lab study with a palette designed for regular expressions (above). Participants were asked to write regular expressions, in the Java programming language, in response to several prompts. The control group was not given access to Graphite, but was otherwise free to use online resources. The treatment group was given a short tutorial of the Graphite system using a palette for the Color class as an example. The fact that a palette existed that was relevant to the tasks they would be asked to perform was mentioned, but the palette itself was not explicitly demonstrated, and its use was not required.

Although our present sample size of 7 participants renders quantitative comparison difficult, several qualitative observations support the view that Graphite was helpful for users in this task. In particular, we found that the participants in the control group were facing difficulties that the palette was designed to address, including difficulty with the factory pattern used by Java regular expressions and the requirement that backslashes be doubly-escaped. We also observed that few subjects wrote adequate testing code to ensure that their regular expressions were correct. Participants in the treatment group, on the other hand, had few or no difficulties with these issues and wrote tests more frequently, due to the support for testing in the palette interface. Feedback from these participants, as well as participants in the control group who were shown the regular expression palette following the experiment, was also positive.

%
%\section{Future Work}
%
%We currently have a list of future works as following:
%
%\begin{itemize}
%	\item Implementing Graphite plug-in for different IDEs
%	\item Improving the Graphite API for Javascript
%	\item Implementing Keyword invocation model
%	\item Evaluating Graphite tool more thoroughly.
%\end{itemize}
%
%
